---
title: "mc_stability_analysis"
output: html_document
date: "2024-02-12"
editor_options: 
  chunk_output_type: console
---

# general package

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("purrr")
library("mapplots")
library("ggplot2")
library("gridExtra")
library("igraph")
library(igraphdata)
library("ggplot2")
library(grid)
library(cowplot)
library(ggtext)
library(zoo)


```


# edm package

```{r}

library(vegan)
library(rEDM) # Some EDM analyses (e.g., simplex projection and CCM) were conducted in rEDM package ver. 1.2.3
packageVersion("rEDM")
library(doParallel)
library(parallel)
library(foreach)
library(Kendall)
library(MASS)
library(dplyr)
library(glmnet)

```


# set directory
```{r}

# set directory
dr <- "/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc"

```


# code settings
##(check if this code is needed)
```{r}

seed <- 49563
set.seed(seed)
SaveFile <- T #T/F for saving files

```


# source Demo_MDR_function.R, mc_mdr_smap.Rmd

```{r}

source("/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc/script/Demo_MDR_function.R")

```


# load ccm.fast.demo.seq.1 from mc_mdr_smap.Rmd

```{r}

ccm.fast.demo.seq.1 <- function(ds,Epair=T,cri='rmse',Emax=10){
  if(cri=='rho'){jfun <- match.fun('which.max')}else{jfun <- match.fun('which.min')}
  ds <- as.matrix(apply(ds,2,scale))
  np <- nrow(ds) # time series length
  ns <- ncol(ds) # number of nodes
  #lib.s <- c(seq(10,nrow(ds),10),nrow(ds)) # sequence of library size
  lib.s <- c(seq(10,nrow(ds),1),nrow(ds))
  crirho <- qt(0.95,np-1)/(np-2+qt(0.95,np-1)^2) # critical values with alpha=0.05
  ccm.rho <- ccm.sig <- matrix(0,ns,ns)
  for(i in 1:ns){
    t.begin <- proc.time()
    for(j in 1:ns){
      # select the optimal E for CCM based on hindcast at time= t-1 (tp=-1)
      ccm.E <- NULL  
      for(E.t in 2:Emax){
        ccm.E <- rbind(ccm.E,ccm(cbind(x=ds[,i],y=ds[,j]), E = E.t, tp=-1,
                              lib_column = "x", target_column = "y", 
                              lib_sizes = nrow(ds),  random_libs =F))
      }
      Eop <- ccm.E[jfun(ccm.E[,cri]),'E'] # The optimal E for the cross-mapping from node i to node j 
      
      
      # Perform CCM at time t (tp=0)      
      ccm.out <- ccm(cbind(x=ds[,i],y=ds[,j]), E = Eop, tp=0, 
                  lib_column = "x", target_column = "y", 
                  lib_sizes = lib.s,  random_libs =F)
      # aggregate the results with respect to each library size
      ccm.seq <- aggregate(ccm.out[,'rho'],list(ccm.out[,'lib_size']),mean,na.rm=T)
      ccm.seq <- ccm.seq[!(is.na(ccm.seq[,2])|is.infinite(ccm.seq[,2])),]
      ccm.seq[ccm.seq[,2]<0,2] <- 0
      termrho <- ccm.seq[nrow(ccm.seq),2]  # rho at the maximal library size (terminal rho)
      if(nrow(ccm.seq)>=3){
        kend <- MannKendall(ccm.seq[,2]);  # Kendall's tau test for mononotic increase
        # Causation is significant only if both (1) Kendall's tau and (2) terminal rho are significantly larger than zero
        ccm.sig[i,j] <- (kend$tau[1]>0)*(kend$sl[1]<0.05)*(termrho>crirho) # ccm.sig records the significance of each CCM
      }else{ccm.sig[i,j] <- 0}
      ccm.rho[i,j] <- termrho                                              # ccm.rho records the terminal rho
    }#end of j
    time.used <- proc.time() - t.begin 
    cat("variable", i, "ccm completed:", time.used[3],"sec\n")
  }#end of i
  return(list(ccm.rho=ccm.rho,ccm.sig=ccm.sig))  
}

```


# Read data

```{r}

# read [al.scale] CPUE per age group for all NS
al.scale

# read [nl.scale] CPUE of total population for all NS
nl.scale

# read jacobian matrix data
file.name.jcof <- paste0(dr, sep = "/","output/6age_all_ns/jcof_1809.csv")

(ja <- read.csv(file.name.jcof))


```
#------[fig 2] 

# use [ja] to estimate stability, save in [sta.dd] 

```{r}

# read & clean [ja]
ja <- as.tibble(ja)

ja <- ja %>% 
  dplyr::select (-j0) %>% 
  filter(Insample ==1) %>% 
  dplyr::select(-Insample) 

# run eigen analysis 
list_result <- list()
list_eigen_position <- list()

for(i in 1:39){
  dd <- ja %>% filter(time == i)
  m <- dd[,3:8]
  # clark's code
  ei.t <- eigen(m)
  ind <- which.max(abs(ei.t$values))
  dev <- abs(ei.t$values[ind]) # add abs() to extract
  list_result[[i]] <- dev
  list_eigen_position[[i]] <- ind 
  # to check whether the dominant is always at the first position
}


# unlist the results
sta.dd <- tibble(Year = seq(from = 1982, to = 2020, 1), Abs.dev = unlist(list_result))

```


# plot abundance log CPUE
```{r}

# set theme_specific [theme_s]
theme_s <- scale_x_continuous(breaks = seq(1980, 2030, by = 5))

# plot abundance over time
age.year.plot <- age.year %>% 
  group_by(Year) %>% 
  mutate (CPUE_all = sum(CPUE_all_subarea))

(abundance.plot <- age.year.plot %>% 
  ggplot()+
  geom_line(aes(x = Year, y =
                  log(CPUE_all_subarea), 
                color = Age), size = 0.5)+
  geom_line(aes(x = Year, 
                y = log(CPUE_all)), size = 1.5)+
  theme+
  ylab("Abundance (log CPUE)")+
  theme_bw() +
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = 'none'
        )+ theme_s+
    scale_color_manual(values=c("#c7e9b4","#7fcdbb","#41b6c4", "#1d91c0", "#225ea8","#0c2c84"))+
    ylim(2,10))
  #+facet_grid(vars(Age))


```


#[fig1a] plot abundance not log transformed
```{r}
# plot abundance over time
age.year.plot <- age.year %>% 
  group_by(Year) %>% 
  mutate (CPUE_all = sum(CPUE_all_subarea))

(abundance.plot.not.log <- age.year.plot %>% 
  ggplot()+
  geom_line(aes(x = Year, y =
                  (CPUE_all_subarea/1000), 
                color = Age), size = 1)+
  geom_line(aes(x = Year, 
                y = (CPUE_all/1000)), size = 1.5)+
  theme+
  ylab("Abundance (thousand CPUE)")+
  theme_bw() +
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = 'none'
        )+ theme_s+
    scale_color_manual(values=c("#c7e9b4","#7fcdbb","#41b6c4", "#1d91c0", "#225ea8","#0c2c84")))


```



#[fig1a] plot abundance not log transformed diff color
```{r}
# plot abundance over time
age.year.plot <- age.year %>% 
  group_by(Year) %>% 
  mutate (CPUE_all = sum(CPUE_all_subarea))

(abundance.plot.not.log.diff.color <- age.year.plot %>% 
  ggplot()+
  geom_line(aes(x = Year, y =
                  (CPUE_all_subarea/1000), 
                color = Age), size = 1)+
  geom_line(aes(x = Year, 
                y = (CPUE_all/1000)), size = 1.5)+
  theme+
  ylab("Abundance (thousand CPUE)")+
  theme_bw() +
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = 'none'
        )+ theme_s+
 scale_colour_brewer(palette="Paired"))

```


#[fig1a] plot abundance legend diff color
```{r}

# plot abundance over time
age.year.plot.legend <- age.year %>% 
  group_by(Year) %>% 
  mutate (CPUE_all = sum(CPUE_all_subarea))

(abundance.plot.not.log.diff.color <- age.year.plot %>% 
  ggplot()+
  geom_line(aes(x = Year, y =
                  (CPUE_all_subarea/1000), 
                color = Age), size = 1)+
  geom_line(aes(x = Year, 
                y = (CPUE_all/1000)), size = 1.5)+
  theme+
  ylab("Abundance (thousand CPUE)")+
  theme(legend.position = "top", legend.direction = "horizontal")+
  # theme_bw() +
  # theme(axis.line = element_line(),
  #       panel.grid.major = element_blank(),
  #       panel.grid.minor = element_blank(),
  #       panel.background = element_blank(),
  #       panel.border = element_blank(),
  #       legend.position = 'none'
  #       )+ theme_s+
 scale_colour_brewer(palette="Paired"))

```


# calculate distance to 1 & rolling mean of stability 
```{r}

# Calculate distance to 1 & absolute distance

sta.dd <- sta.dd %>% 
  mutate(distance_1 = Abs.dev-1,
         distance_1_abs =abs(distance_1))


# calculate rolling mean of distance

sta.dd <- sta.dd %>%
  mutate(distance_1_roll_5 = rollmean(distance_1, k=5, fill=NA, align='right')) %>% 
  mutate(distance_1_abs_roll_5 = rollmean(distance_1_abs, k=5, fill=NA, align='right'))

  
# define positive or negative eigenvalue

sta.dd <- sta.dd %>% 
   mutate(dev.sign = 
            replace(Abs.dev, Abs.dev <1, "<1")) %>% 
   mutate(dev.sign = replace(dev.sign, Abs.dev >1, ">1"))
  

# Calculate rolling mean

sta.dd <- sta.dd %>%
  mutate(roll_3 = rollmean(Abs.dev, k=3, fill=NA, align='right')) %>% 
  mutate(roll_4 = rollmean(Abs.dev, k=4, fill=NA, align='right')) %>% 
  mutate(roll_5 = rollmean(Abs.dev, k=5, fill=NA, align='right'))
  


```


# [fig1c] plot stability over time
```{r}

(stability.plot <- ggplot(sta.dd)+ 
  geom_line(size = 1,aes(x = Year, y = Abs.dev))+
  xlab("Year")+
  ylab("Population stability")+
  theme_bw()+
  theme_s+
  geom_hline(yintercept=1, linetype="dashed")+
   scale_y_continuous(
    breaks = seq(from =0.4, to = 1.6,
                       by = 0.2),
    limits = c(0.2, 1.6))+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = c(0.9, 0.7))+
   geom_line(size = 1, color = 'red',
             aes(x = Year, y = roll_5))+
   geom_vline(xintercept=2000, linetype="dashed"))
  


# bind two plots
arranged.plot <- grid.arrange(abundance.plot, stability.plot)

grid.text("a)", x = unit(0.01, "npc"), y = unit(0.9, "npc") + unit(8, "mm"), just = "left") # add (a)
grid.text("b)", x = unit(0.01, "npc"), y = unit(0.4, "npc") + unit(8, "mm"), just = "left") # add (b)

```

# plot distance to 1
```{r}

(distance.plot <- ggplot(sta.dd,
                         aes(x = Year, y = distance_1))+ 
  geom_col()+
  xlab("Year")+
  ylab("distance to 1")+
  theme_bw()+
  theme_s+
  geom_hline(yintercept = 0, linetype="dashed")+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = c(0.9, 0.7))+
   geom_line(size = 1, color = 'red',
             aes(x = Year, y = distance_1_roll_5))+
    geom_vline(xintercept=2000, linetype="dashed"))


(distance.absolute.plot <- 
    ggplot(sta.dd,
           aes(x = Year, y = distance_1_abs))+ 
  geom_col(aes(fill = dev.sign))+
  xlab("Year")+
  ylab("absolute distance to 1")+
  theme_bw()+
  theme_s+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = c(0.9, 0.7))+
     geom_line(size = 1, color = 'black',
             aes(x = Year, y = distance_1_abs_roll_5))+
     geom_vline(xintercept=2000, linetype="dashed"))
  

# plot stability, distance, and absolute distance

(all3.plot <- grid.arrange(stability.plot,
                       distance.plot,
                       distance.absolute.plot))

```


# [fig1a] abundance plot with legend
```{r}

(abundance.legend.plot <- age.year.plot %>% 
  ggplot()+
  geom_line(aes(x = Year, y =
                  log(CPUE_all_subarea), 
                color = Age))+
  geom_line(aes(x = Year, 
                y = log(CPUE_all)),
            size = 1.2)+
  theme+
  ylab("Abundance \n (log CPUE)")+
  theme_bw() +
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = 'top',
        )+ theme_s+
    guides(color = guide_legend(nrow = 1))+
   scale_color_manual(values=c("#c7e9b4","#7fcdbb","#41b6c4", "#1d91c0", "#225ea8","#0c2c84"))) # legend 1 line


```


# plot by age (not finished yet)
```{r}

# set theme_specific [theme_s]
theme_s <- scale_x_continuous(breaks = seq(1980, 2030, by = 5))

# bind abundance by age and total abundance


(abundance.plot <- age.year.plot %>% 
  ggplot()+
  geom_line(aes(x = Year, y =
                  log(CPUE_all_subarea), 
                color = Age), size = 1)+
  geom_line(aes(x = Year, 
                y = log(CPUE_all)), size = 1.4)+
  theme+
  ylab("Abundance \n (log CPUE)")+
  theme_bw() +
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = 'none'
        )+ theme_s+
    scale_color_manual(values=c("#c7e9b4","#7fcdbb","#41b6c4", "#1d91c0", "#225ea8","#0c2c84"))+
    facet_grid(vars(Age)))


age.year 

dd.year %>% 
  mutate(Age = '0_total') 

```


# plot age structure per 10 years
```{r}

# prepare structure.dd
structure.dd <- age.year %>%
  mutate(period = 
           replace(Year,
                  Year <= 1990, 
                   "1982-1990")) %>% 
  mutate(period = 
           replace(period,
                  Year <= 2000 & Year >=1991, 
                   "1991-2000")) %>% 
  mutate(period = 
           replace(period,
                  Year <= 2010 & Year >=2001, 
                   "2001-2010")) %>% 
  mutate(period = 
           replace(period,
                   Year >=2011, 
                   "2011-2021")) %>% 
  group_by(Age,period) %>% 
  summarise(cpue = sum(CPUE_all_subarea))

  

(structure.plot <- ggplot(data= structure.dd, 
                         aes(x=Age, y=cpue/1000, fill = Age)) +
  geom_bar(stat="identity")+
  facet_wrap(vars(period), ncol = 4)+
      scale_fill_manual(values=c("#c7e9b4","#7fcdbb","#41b6c4", "#1d91c0", "#225ea8","#0c2c84"))+
  theme_bw() +
  theme(
        axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        #panel.border = element_blank(),
        legend.position = 'none',
        # plot.margin = margin(t = 0, r = 10, 
        #                      b = 0, l = 0, unit = "pt")
        )+
  ylab('Abundance (thousand CPUE)'))
  
  
  
```

# [fig1b] plot age structure per 20 years
```{r}

# prepare structure.dd
structure.dd.20yr <- age.year %>%
  mutate(period = 
           replace(Year,
                  Year <= 2000, 
                   "1982-2000")) %>% 
  mutate(period = 
           replace(period,
                   Year >=2001, 
                   "2001-2021")) %>% 
  group_by(Age,period) %>% 
  summarise(cpue = sum(CPUE_all_subarea))

  

(structure.plot.20yr <- ggplot(data= structure.dd.20yr, 
                         aes(x=Age, y=cpue/1000, fill = Age)) +
  geom_bar(stat="identity")+
  facet_wrap(vars(period), ncol = 2)+
      scale_fill_manual(values=c("#c7e9b4","#7fcdbb","#41b6c4", "#1d91c0", "#225ea8","#0c2c84"))+
  theme_bw() +
  theme(
        axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        #panel.border = element_blank(),
        legend.position = 'none',
        # plot.margin = margin(t = 0, r = 10, 
        #                      b = 0, l = 0, unit = "pt")
        )+
  ylab('Abundance (thousand CPUE)')+
  ylim(0,80))
  
  
  
```


# [fig1b] plot age structure per 20 years: diff color
```{r}

# prepare structure.dd
structure.dd.20yr <- age.year %>%
  mutate(period = 
           replace(Year,
                  Year <= 2000, 
                   "1982-2000")) %>% 
  mutate(period = 
           replace(period,
                   Year >=2001, 
                   "2001-2021")) %>% 
  group_by(Age,period) %>% 
  summarise(cpue = sum(CPUE_all_subarea))

  

(structure.plot.20yr.diff.color <- ggplot(data= structure.dd.20yr, 
                         aes(x=Age, y=cpue/1000, fill = Age)) +
  geom_bar(stat="identity")+
  facet_wrap(vars(period), ncol = 2)+
      # scale_fill_manual(values=c("#c7e9b4","#7fcdbb","#41b6c4", "#1d91c0", "#225ea8","#0c2c84"))+
    scale_fill_brewer(palette="Paired")+
  theme_bw() +
  theme(
        axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        #panel.border = element_blank(),
        legend.position = 'none',
        # plot.margin = margin(t = 0, r = 10, 
        #                      b = 0, l = 0, unit = "pt")
        )+
  ylab('Abundance (thousand CPUE)')+
  ylim(0,80))
  
  
  
```


# bind 3 plots: abundance+ age structure+ stability
```{r}
# bind 3 plots
arranged.plot <- grid.arrange(abundance.plot.not.log, structure.plot.20yr, stability.plot)

grid.text("(a)", x = unit(0.01, "npc"), y = unit(0.9, "npc") + unit(8, "mm"), just = "left") # add (a)
grid.text("(b)", x = unit(0.01, "npc"), y = unit(0.6, "npc") + unit(8, "mm"), just = "left") # add (b)
grid.text("*(c)", x = unit(0.01, "npc"), y = unit(0.3, "npc") + unit(8, "mm"), just = "left") # add (c)
```


# bind 3 plots: abundance+ age structure+ stability:diff color
```{r}
# bind 3 plots
arranged.plot <- grid.arrange(abundance.plot.not.log.diff.color, structure.plot.20yr.diff.color, stability.plot)

grid.text("(a)", x = unit(0.01, "npc"), y = unit(0.9, "npc") + unit(8, "mm"), just = "left") # add (a)
grid.text("(b)", x = unit(0.01, "npc"), y = unit(0.6, "npc") + unit(8, "mm"), just = "left") # add (b)
grid.text("(c)", x = unit(0.01, "npc"), y = unit(0.3, "npc") + unit(8, "mm"), just = "left") # add (c)

```

# calculate statistics to put in the graph
```{r}

# proportion of age to total abundance
(structure.dd.20yr %>% 
  group_by(period) %>% 
  mutate(total_cpue = sum(cpue)) %>% 
  ungroup() %>% 
  mutate(proportion = round ((cpue/total_cpue)*100,0)))

# mean and sd of population stability
(sta.dd %>% 
  mutate(period = 
           replace(Year,
                  Year <= 2000, 
                   "1982-2000")) %>% 
  mutate(period = 
           replace(period,
                   Year >=2001, 
                   "2001-2021")) %>% 
  group_by(period) %>% 
  summarise(mean_dev = round(mean(Abs.dev),2),
            sd_dev = round(sd(Abs.dev),2)))
  

# check difference before and after 2000 for age 1 and 2
20724.909/54825.356	 # 37% 
100-37 # dropped 63%
20705.683/58602.759	 # 35%
100-35 # dropped 65%


```


# calculate mean abundance 1983-2000, 2001-2021
```{r}

# mean abundance
dd.year %>% 
  mutate(phase = 
           replace(Year,
                  Year <= 2000, 
                   "before_2000")) %>% 
  mutate(phase = 
           replace(phase,
                  Year>=2001, 
                   "after_2000")) %>% 
  group_by(phase) %>% 
  summarise(cpue = mean(CPUE_all_subarea_age))
  
2847/7524 # 0.37 proportion of total abundance before 2000/after 2000


# proportion of age 1 and 2 
age.year %>% 
  mutate(phase = 
           replace(Year,
                  Year <= 2000, 
                   "before_2000")) %>% 
  mutate(phase = 
           replace(phase,
                  Year>=2001, 
                   "after_2000")) %>% 
  mutate(age_2groups =
           replace(Age,
                   Age %in% c('1','2'), 'age1_2')) %>% 
  mutate(age_2groups =
           replace(age_2groups,
                   Age %in% c('3','4',
                              '5','6'), 'age3_6')) %>% 
  group_by(age_2groups, phase) %>% 
  # here we get cpue of age 1-2 & 3-6 before 2000 & after 2000
  summarise(cupe = sum(CPUE_all_subarea)) 


113428/(113428+29543) #79 %  age 1-2/total age before 2000
41430/(41430+18375) #69 % age 1-2/total age after 2000


```




#----------[fig 3] interactions between age groups

# transform ja to long form <ja.dd>

Note that for jacobian matrix output [ja], columns "A1","A2","A3","A4","A5","A6" are explanatory variables. The second column named "variable" is the responsible variable (predicted age). This direction is the same as the jacobian matrix.

output 7 x 10
```{r}

ja.dd <- ja %>%
 pivot_longer(
   cols = c("A1","A2","A3","A4","A5","A6"),
   names_to = "explanatory_age",
   values_to = "jacobian",
   values_drop_na = TRUE) %>% 
  mutate (response_age = 
            paste("A", sep = "", variable)) %>% 
  #select(-variable) %>% 
  relocate(response_age, .before = jacobian) %>% 
  mutate(year = 1981+time)
  
```


# compute Jii-1 and create <ja.dd.1>

```{r}

ja.dd.1 <- ja.dd %>% 
   mutate(jacobian = ifelse(explanatory_age == response_age,jacobian-1 , jacobian))
  
```


# check why a3 has almost constant interactions over time
```{r}

ja.dd.1 %>% filter(response_age =='A3')
ja %>% filter(variable =='3')

```


# [Fig S2] plot all interactions over time
note: use ja.dd or ja.dd.1 with jii-1 
```{r}

# plot each age as predicted variable

fun_plot_interaction <- function(filter_res_age, title){
  dd <- ja.dd.1 %>% filter(response_age  == filter_res_age) 
  ggplot(dd,aes(x = year, y = jacobian, color = explanatory_age))+ 
  geom_point()+
  geom_line()+theme_bw()+ 
    ylim(-2.3, 2.3)+
    ylab('Interaction strength')+ xlab('Year')+
    ggtitle(paste0(title, 
                   sep = " ", 
                   "as predictor"))+
    scale_x_continuous(breaks = seq(1980, 2030, by = 5))+
    scale_colour_brewer(palette="Paired")+
    theme_bw() +
      labs( col="explanatory age")+

  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank())
}



p1 <- fun_plot_interaction(filter_res_age = "A1", title = "age-1 group")
p2 <- fun_plot_interaction(filter_res_age = "A2", title = "age-2 group")
p3 <- fun_plot_interaction(filter_res_age = "A3", title = "age-3 group")
p4 <- fun_plot_interaction(filter_res_age = "A4", title = "age-4 group")
p5 <- fun_plot_interaction(filter_res_age = "A5", title = "age-5 group")
p6 <- fun_plot_interaction(filter_res_age = "A6", title = "age-6+ group")

grid.arrange(p1,p2,p3,p4,p5,p6, ncol=2) # output 7 x 10


```


# Plot self-regulation
```{r}

fun_plot_links <- 
  function(filter_res, filter_exp, title, title2){
  dd <- ja.dd.1 %>% 
    filter(response_age  == filter_res,
           explanatory_age == filter_exp)
  
  ggplot(dd,aes(x = year, y = jacobian))+ 
  geom_point()+
  geom_line()+ xlab('Year')+
    ylim(-2.3, 2.3)+ ylab("Interaction strength")+
    geom_hline(yintercept=0, linetype="dashed")+
    ggtitle(paste0(title, sep = " ", 
                   title2))+
    scale_x_continuous(breaks = seq(1980, 2030, by = 5))+
    theme_bw() +
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank())
    
}


# plot self-regulation strengths
p1 <- fun_plot_links (filter_res = "A1", filter_exp = "A1",
                      title = "Self-regulation of", 
                      title2 = " age-1 group")
p2 <- fun_plot_links (filter_res = "A2", filter_exp = "A2",
                      title = "Self-regulation of", 
                      title2 = "age-2 group")
p3 <- fun_plot_links (filter_res = "A3", filter_exp = "A3",
                      title = "Self-regulation of", 
                      title2 = "age-3 group")
p4 <- fun_plot_links (filter_res = "A4", filter_exp = "A4",
                      title = "Self-regulation of", 
                      title2 = "age-4 group")
p5 <- fun_plot_links (filter_res = "A5", filter_exp = "A5",
                      title = "Self-regulation of", 
                      title2 = "age-5 group")
p6 <- fun_plot_links (filter_res = "A6", filter_exp = "A6",
                      title = "Self-regulation of", 
                      title2 = "age-6+ group")

grid.arrange(p1,p2,p3,p4,p5,p6, ncol=2) # output 7 x 10

```

# Plot growth from one age to the next
```{r}

p1 <- fun_plot_links (filter_res = "A2", filter_exp = "A1",
                      title = "age 1", 
                      title2 = "grow to age 2")
p2 <- fun_plot_links (filter_res = "A3", filter_exp = "A2",
                      title = "age 2", 
                      title2 = "grow to age 3")
p3 <- fun_plot_links (filter_res = "A4", filter_exp = "A3",
                      title = "age 3", 
                      title2 = "grow to age 4")
p4 <- fun_plot_links (filter_res = "A5", filter_exp = "A4",
                      title = "age 4", 
                      title2 = "grow to age 5")
p5 <- fun_plot_links (filter_res = "A6", filter_exp = "A5",
                      title = "age 5", 
                      title2 = "grow to age 6")

grid.arrange(p1,p2,p3,p4,p5, ncol=2) # output 7 x 10

```


# take mean jacobian over time [ja.dd.mean] and plot
note: use ja.dd or ja.dd.1 with jii-1
```{r}

# calculate mean jacobian over time
ja.dd.mean <- ja.dd %>% group_by(explanatory_age, response_age) %>%
  summarise(mean_ja = mean(jacobian)) 

# plot
ggplot(ja.dd.mean,aes(y = mean_ja, x = explanatory_age))+ 
  geom_boxplot()+
  facet_wrap("response_age")+theme_bw()


```


# assign jacobian by sign & strength [ja.dd.mean.cat]

categorise to positive/negative with weak, medium, strong links

```{r}

ja.dd.mean.cat <- ja.dd.mean %>% 
  # remove 3 edges which mean jacobian = 0
  filter (mean_ja != 0) %>% 
  # create mean_jacobian_sign to classify into positive/negative edges
  mutate(mean_ja_sign = 
           ifelse((mean_ja > 0), 
                  "positive", "negative")) %>% 
  # create mean_jacobian_strength for positive values
  mutate(mean_ja_strength = 
           replace(mean_ja,
                   between(mean_ja, 0,0.2), 
                   "weak")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, 0.2,0.4), 
                   "medium")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                  mean_ja > 0.4, 
                   "strong")) %>% 
  # create mean_jacobian_strength for negative values
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, -0.2,0), 
                   "weak")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, -0.4,-0.2), 
                   "medium")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                  mean_ja < - 0.4, 
                   "strong")) 

```


# assign jacobian by sign & strength before after 2000 [ja.dd.mean.cat.year]

```{r}

# take mean jacobain before & after year 2000

(ja.dd.mean.period <- ja.dd %>% 
  mutate(period = 
           replace(year,
                  year <= 2000, '1982-2000')) %>% 
  mutate(period = 
           replace(period,
                  year > 2000, 
                   "2000-2020")) %>% 
  group_by(explanatory_age, response_age, period) %>%
  summarise(mean_ja = mean(jacobian))) 


# assign jacobian by sign and strength

ja.dd.mean.cat.period <- ja.dd.mean.period %>% 
  # remove 3 edges which mean jacobian = 0
  filter (mean_ja != 0) %>% 
  # create mean_jacobian_sign to classify into positive/negative edges
  mutate(mean_ja_sign = 
           ifelse((mean_ja > 0), 
                  "positive", "negative")) %>% 
  # create mean_jacobian_strength for positive values
  mutate(mean_ja_strength = 
           replace(mean_ja,
                   between(mean_ja, 0,0.2), 
                   "weak")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, 0.2,0.4), 
                   "medium")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                  mean_ja > 0.4, 
                   "strong")) %>% 
  # create mean_jacobian_strength for negative values
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, -0.2,0), 
                   "weak")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, -0.4,-0.2), 
                   "medium")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                  mean_ja < - 0.4, 
                   "strong")) %>%
  # sort tibble by period
  arrange(period) %>% 
  # relocate columns
  relocate(period, .before = 'explanatory_age') %>%
  relocate(mean_ja, .after = 'mean_ja_strength')

ja.dd.mean.cat.period


```

# jii-1: assign jacobian by sign & strength before after 2000 [ja.dd.mean.cat.year]
note: category by 0.3 for 4 categories
```{r}

# take mean jacobain before & after year 2000

(ja.dd.mean.period.1<- ja.dd.1%>% 
  mutate(period = 
           replace(year,
                  year <= 2000, '1982-2000')) %>% 
  mutate(period = 
           replace(period,
                  year > 2000, 
                   "2000-2020")) %>% 
  group_by(explanatory_age, response_age, period) %>%
  summarise(mean_ja = mean(jacobian))) 


# assign jacobian by sign and strength

ja.dd.mean.cat.period.1 <- ja.dd.mean.period.1 %>% 
  group_by(explanatory_age, response_age) %>% 
  # remove 3 edges which mean jacobian = 0
  filter (mean_ja != 0) %>% 
  # create mean_jacobian_sign to classify into positive/negative edges
  mutate(mean_ja_sign = 
           ifelse((mean_ja > 0), 
                  "positive", "negative")) %>% 
  # create mean_jacobian_strength for positive values
  mutate(mean_ja_strength = 
           replace(mean_ja,
                   between(mean_ja, 0,0.3), 
                   "weak")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, 0.3,0.6), 
                   "medium")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, 0.6,0.9), 
                   "strong")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                  mean_ja > 0.9, 
                   "very strong")) %>% 
  # create mean_jacobian_strength for negative values
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, -0.3,0), 
                   "weak")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, -0.6,-0.3), 
                   "medium")) %>% 
   mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, -0.9,-0.6), 
                   "strong")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                  mean_ja < - 0.9, 
                   "very strong")) %>%
  # sort tibble by period
  arrange(period) %>% 
  # relocate columns
  relocate(period, .before = 'explanatory_age') %>%
  relocate(mean_ja, .after = 'mean_ja_strength')

ja.dd.mean.cat.period.1

# # check min & max mean jacobian value
# ja.dd.mean.cat.period.1 %>% 
#   arrange(mean_ja) # -1.2 - 0.7

```


#--------identify most vulnerable group eigenvector ranking (Medeiros et al 2023)

# code from Medeiros 2023 <forecast_smap_empirical_time_series>
- Based on Medeiros 2023, leading eivenvalue is the largest real part of the eigenvalue (quote Medeiros 2023 Box 2: vi is the real part of the ith eigenvector of J, 𝜆i is the real part of the ith eigenvalue). 

- Based on Medeiros 2023, ranking of species is the order of absolute value of element of the dominant eigenvector (quote Medeiros 2023: we compute the alignment of species i with v1 as the absolute value of its ith element (∣ v1i ∣), where ‖ v1 ‖ = 1. We define the order of ∣ v1i ∣ values across species as the eigenvector ranking and use it to predict the order of ⟨si⟩)

```{r, eval = FALSE}

#-----from Medeiros 2023 <forecast_smap_empirical_time_series>

## leading eigenvalue
smap_eigenvalue[i] <- max(Re(eigen_dec$values))

## leading eigenvector
order_values <- order(Re(eigen_dec$values), decreasing = TRUE)
smap_eigenvector[[i]] <- Re(eigen_dec$vectors)[ , order_values[1]]


#----from Medeiros 2023 <jacobian_sensitivities_synthetic_time_series>

## decreasing order of eigenvalues
values_order <- order(Re(eigen_J_list[[i]]$values), decreasing = TRUE)

## extracting ordered eigenvalues
values[[i]] <- Re(eigen_J_list[[i]]$values)[values_order]

## extracting leading eigenvector 
vectors[[i]] <- Re(eigen_J_list[[i]]$vectors)[ , values_order[1]]

## eigenvector alignments
### the denominator is possibly the sum of square of all element of the dominant eigenvector, it is for standardization.
eigen_alignments[[i]] <- abs(vectors[[i]] / sqrt(sum(vectors[[i]]^2))) 

```


# calculate eigenvector alignment, result in [eigen.alignment]
Based on Medeiros 2023, I took the mean of the raking over time to define the final rank

```{r}

dd <- ja %>% filter(time == 1)
m <- dd[,3:8]

#----snippet: based on Meidros's code
## order eigenvalue
order_values <- order(Re(eigen(m)$values), decreasing = TRUE)
## extract dominant eigenvector
smap_eigenvector <- Re(eigen(m)$vectors)[order_values[1],]
## standardize and take absolute value for each element
eigen_alignments <- abs(smap_eigenvector / sqrt(sum(smap_eigenvector^2)))


# for loop moving forward one point at a time
list_eigen_alignment <- list()

for(i in 1:39){
  dd <- ja %>% filter(time == i)
  m <- dd[,3:8]
  ## order eigenvalue
  order_values <- order(Re(eigen(m)$values), decreasing = TRUE)
  ## extract dominant eigenvector
  smap_eigenvector <- Re(eigen(m)$vectors)[order_values[1],]
  ## standardize and take absolute value for each element
  eigen_alignments <- abs(smap_eigenvector / sqrt(sum(smap_eigenvector^2)))
  # store result
  list_eigen_alignment [[i]] <- eigen_alignments
  }

# unlist the results
eigen.alignment <- tibble(
  age = rep(seq(from = 1, to = 6, 1), times = 39), 
  year = rep(seq(from = 1982, to = 2020, 1), each = 6),
  eigen_alignment = unlist(list_eigen_alignment)) %>% 
  mutate_at("age", as.factor)

```


# plot eigen alignment per age over time
```{r}

eigen.alignment

age_names <- c(
                    `1` = "age 1",
                    `2` = "age 2",
                    `3` = "age 3",
                    `4` = "age 4",
                    `5` = "age 5",
                    `6` = "age 6"
                    )

eigen.age <- ggplot(eigen.alignment,
       aes(x = year, y = eigen_alignment))+ 
  geom_point()+
  geom_line()+
  facet_wrap(vars(age), 
             labeller = as_labeller(age_names),
             strip.position="left",
             ncol = 1)+
  xlab("Year")+
  ylab("")+
  theme_bw()+
  theme(
        axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  # set lable position and rotate
  theme(strip.background = element_blank(),
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0)
  )+
  scale_y_continuous(breaks = seq(0, 1, by = 0.3))+
  # rotate y lable
  #theme(axis.title.y = element_text(angle = 0, vjust = 0.5))+
  ggtitle("Relative contribution to population stability")

  
eigen.age        


```


# plot younger (age 1+2+3) and older (age 3+5+6)
```{r}

# 1 create [eigen.alignment.young.old]
eigen.alignment.young.old <- eigen.alignment %>% 
  mutate(older_younger = 
           ifelse((age %in% c(4,5,6)),
                  'older','younger')) %>% 
  group_by(year,older_younger) %>% 
  summarise(sum_alignment = sum(eigen_alignment))

# 2 plot, color as younger, older
ggplot(eigen.alignment.young.old,
       aes(x = year, y = sum_alignment,
           color = older_younger))+ 
  geom_point()+
  geom_line()+
  xlab("Year")+
  ylab("contribution to population stability")+
  theme_bw()


eigen.alignment %>% 
  group_by(age) %>% 
  summarise(mean_eigen = mean(eigen_alignment))

```


# eigenvector violin plot
```{r}


# violin plot
eigen.violin <- ggplot(eigen.alignment, aes(x=age, y=eigen_alignment, fill = age)) + # fill=name allow to automatically dedicate a color for each group
  geom_violin(width=1.6, color="grey")+
  geom_boxplot(width = 0.3) +
  geom_point(size=1,shape=21,
             position=position_jitterdodge(1)) +
  theme_bw()+
  theme(legend.position="none",
        axis.title.y = element_text(angle = 0, vjust = 0.5))+
  #ylab("Contribution to \npopulation stability") +
  xlab('')+ylab('')+
  #ggtitle("alignment values from each of 39 years")+
  scale_fill_brewer(palette="Paired")+
  scale_x_discrete(
    labels= c('age 1', 
              'age 2',
              'age 3', 
              'age 4',
              'age 5',
              'age 6'))+
  ggtitle("Relative contribution to population stability over time")
  
eigen.violin

```

# [fig 4] eigenvector violin before & after 2000
```{r}

eigen.alignment.period <- eigen.alignment %>% 
  mutate(period = 
           replace(year,
                  year <= 2000, 
                   "1982-2000")) %>% 
  mutate(period = 
           replace(period,
                  year > 2000, 
                   "2001-2020")) 
  
  
ggplot(eigen.alignment.period, 
         aes(x=age, y=eigen_alignment, fill = age)) + 
  # fill=name allow to automatically dedicate a color for each group
  geom_violin(width=1.6, color="grey")+
  geom_boxplot(width = 0.3) +
  geom_point(size=1,shape=21,
             position=position_jitterdodge(1)) +
  facet_wrap(vars(period), ncol = 1)+
  theme_bw()+
  theme(legend.position="none",
        axis.title.y = element_text(angle = 0, vjust = 0.5))+
  xlab('')+ylab('')+
  scale_fill_brewer(palette="Paired")+
  scale_x_discrete(
    labels= c('age 1', 
              'age 2',
              'age 3', 
              'age 4',
              'age 5',
              'age 6'))+
  ggtitle("Relative contribution to population stability")
  
 

```


# [fig S3] plot eigenvector by ranking
```{r}

# assign rank
(eigen.alignment.period.rank <- 
  eigen.alignment.period %>% 
  group_by(year) %>%
  arrange((eigen_alignment)) %>% 
  mutate(rank = 1:n()) %>% 
  arrange(year)) 
  

# boxplot
ggplot(eigen.alignment.period.rank, 
         aes(x= age, y= rank, fill = age)) + 
  # fill=name allow to automatically dedicate a color for each group
  #geom_violin(width=1.6, color="grey")+
  geom_boxplot(width = 0.3) +
  geom_point(size=1,shape=21,
             position=position_jitterdodge(1)) +
  facet_wrap(vars(period), ncol = 1)+
  theme_bw()+
  theme(legend.position="none",
        axis.title.y = element_text(angle = 0, vjust = 0.5))+
  xlab('')+ylab('')+
  scale_fill_brewer(palette="Paired")+
   scale_y_continuous(breaks = seq(1, 6, by = 1))+
  scale_x_discrete(
    labels= c('age 1', 
              'age 2',
              'age 3', 
              'age 4',
              'age 5',
              'age 6'))+
  ggtitle("Relative contribution to population stability with ranking")+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank())

```




